## 1. Loading the libraries

```{r}
library(tidyverse)
library(janitor)
library(cheapr)
library(maps)
library(plotly)
library(ggrepel)
library(ggplot2)
library(leaflet)
library(sf)
library(tigris)
```

## 2. Ingesting the data

```{r}
setwd("C:/Users/Saima Abbas/Desktop/CAPSTONE DOCS")
MD_counties <-read_csv("Maryland_Counties_Socioeconomic_Characteristics__ACS_5-yr_Estimates_2022_.csv")
MCPS <-read_csv("hs_demographic_other_information.csv")
census_raw <-read_csv("ACSST1Y2024.S1701-2026-05-02T232952.csv")
```

## 3. Cleaning

## Cleaning

#renaming variables

census:

```{r}
census_clean <- census_raw |>
  rename(
    demographic = `Label (Grouping)`,
    poverty_rate = `Montgomery County, Maryland!!Percent below poverty level!!Estimate`
  ) |>
  filter(!is.na(poverty_rate)) |>
  mutate(
    poverty_rate = as.numeric(gsub("%", "", poverty_rate)))
```

removing missing values

```{r}
census_clean <- census_clean |>
  filter(!is.na(poverty_rate)) |>
  na.omit(demographic)
```

reshaping

```{r}
poverty_col <- colnames(census_raw)[str_detect(colnames(census_raw), "Percent")]
```

```{r}
census_clean <- census_raw |>
  rename(poverty_percent = all_of(poverty_col))
```

## MD_counties:

```{r}
MD_counties <- clean_names(MD_counties)
```

```{r}
colnames(MD_counties)
```

```{r}
MD_counties <- MD_counties |>
  rename(
    county = jurisdictions,
    poverty_rate = percent_families_in_poverty,
    median_income = median_household_income,
    unemployment_rate = unemployment_rate
  )
```

```{r}
MD_clean <- MD_counties |>
  filter(!is.na(poverty_rate))
```

```{r}
nrow(MD_clean)
```

## MCPS:

```{r}
MCPS <- clean_names(MCPS)
```

```{r}
MCPS <- MCPS |>
  rename(
    school = school_name,
    grad_rate = graduation_rate,
    dropout_rate = dropout_rate,
    attendance_rate = attendance_rate,
    FARMS_rate = farms
  )
```

```{r}
colnames(MCPS) 
```

```{r}
MCPS_clean <- MCPS |>
  filter(!is.na(grad_rate), !is.na(FARMS_rate))
nrow(MCPS_clean)
```

## 4. summary stats +

# mcps

```{r}
summary(MCPS_clean$grad_rate)
```

```{r}
summary(MCPS_clean$grad_rate)
```

```{r}
summary(MCPS_clean$FARMS_rate)
```

#Mean, Median and standard Deviation

```{r}
MCPS_clean |>
  summarise(
    mean_grad = mean(grad_rate, na.rm = TRUE),
    median_grad = median(grad_rate, na.rm = TRUE),
    sd_grad = sd(grad_rate, na.rm = TRUE),
    
    mean_farms = mean(FARMS_rate, na.rm = TRUE),
    median_farms = median(FARMS_rate, na.rm = TRUE),
    sd_farms = sd(FARMS_rate, na.rm = TRUE)
  )
```

## MD

```{r}
summary(MD_clean$poverty_rate)
```

```{r}
summary(MD_clean$median_income)
```

```{r}
summary(MD_clean)
```

```{r}
MD_clean |>
  summarise(
    mean_poverty = mean(poverty_rate, na.rm = TRUE),
    median_poverty = median(poverty_rate, na.rm = TRUE),
    sd_poverty = sd(poverty_rate, na.rm = TRUE),

    mean_income = mean(median_income, na.rm = TRUE),
    median_income = median(median_income, na.rm = TRUE),
    sd_income = sd(median_income, na.rm = TRUE),

    mean_unemployment = mean(unemployment_rate, na.rm = TRUE),
    median_unemployment = median(unemployment_rate, na.rm = TRUE),
    sd_unemployment = sd(unemployment_rate, na.rm = TRUE))
```

## Census:

```{r}

```

## 5. Visualizations (w questions)

# How does farm participation impact graduation rates?

```{r}
ggplot(MCPS_clean, aes(x = FARMS_rate, y = grad_rate, color = attendance_rate)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(
    title = "FARMS vs Graduation (Colored by Attendance)",
    x = "FARMS Rate (Poverty Proxy)",
    y = "Graduation Rate",
    color = "Attendance Rate"
  ) +
  theme_get()
```

## How does median income and poverty vary across the state?

```{r}
ggplot(MD_clean, aes(x = reorder(county, poverty_rate), 
                     y = poverty_rate, 
                     fill = median_income)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    title = "Poverty Rate by County (Colored by Income)",
    x = "County",
    y = "Poverty Rate (%)",
    fill = "Median Income"
  ) +
  theme_classic()
```

```{r}
MD_clean <- MD_clean |>
  mutate(
    county = gsub(" County", "", county))
```

```{r}
md_map <- counties(state = "MD", cb = TRUE, class = "sf")
```

```{r}
colnames(md_map)
```

```{r}
## counties
md_map <- md_map |>
  mutate(county = gsub(" County", "", NAMELSAD))
```

```{r}
## joining things
md_joined <- md_map |>
  left_join(MD_clean, by = "county")
```

```{r}
## labels
md_joined <- md_joined |>
  mutate(
    county_short = substr(county, 1, 3))
```

maps to pair w this:

```{r}
ggplot(md_joined) +
  geom_sf(aes(fill = `poverty_rate`), color = "white") +
  scale_fill_gradient(low ="lightpink", high = "darkred", na.value = "gray") +
  geom_sf_text(aes(label = county), size = 3) +
  labs(
    title = "Poverty Rate by County (Maryland)",
    fill = "Poverty Rate (%)"
  ) +
  theme_minimal()
```

```{r}
ggplot(md_joined) +
  geom_sf(aes(fill = `median_income`), color = "white") +
  scale_fill_gradient(low ="lightblue", high = "darkblue", na.value = "gray") +
  geom_sf_text(aes(label = county), size = 3) +
  labs(
    title = "Median Income by County (Maryland)",
    fill = "Median Income") +
  theme_minimal()
```

## How does farms correlate to the dropout rate?

```{r}
ggplot(MCPS_clean, aes(x = FARMS_rate, y = grad_rate, color = total)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  scale_color_gradient(low = "lightpink", high = "darkred") +
  labs(
    title = "FARMS vs Dropout (Colored by Graduation Rate)",
    x = "FARMS Rate",
    y = "Dropout Rate",
    color = "Graduation Rate"
  ) +
  theme_minimal()
```

## Does education level relate to income?

```{r}
ggplot(MD_clean, aes(x = bachelors_degree, y = median_income)) +
  geom_point(color = "purple", size = 3) +
  geom_smooth(method = "lm") +
  labs(
    title = "Education vs Income"
  ) +
  theme_minimal()
```

# Does attendance mediate poverty's effect?

```{r}
ggplot(MCPS_clean, aes(x = FARMS_rate, y = attendance_rate)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm") +
  labs(
    title = "Poverty vs Attendance"
  ) +
  theme_minimal()
```

## Which schools perform unexpectedly?

```{r}
ggplot(MCPS_clean, aes(x = FARMS_rate, y = grad_rate, label = school)) +
  geom_point() +
  geom_text(size = 3, check_overlap = TRUE) +
  labs(title = "Outlier Schools")
```

```{r}
ggplot(MD_clean |> arrange(median_income), 
       aes(x = reorder(county, median_income), y = median_income, group = 1)) +
  geom_line() +
  geom_point() +
  coord_flip() +
  labs(
    title = "Income Trend Across Counties"
  )
```

## Do wealthier counties have lower unemployment and poverty?

```{r}
ggplot(MD_clean, aes(x = median_income, y = unemployment_rate, size = poverty_rate)) +
  geom_point(alpha = 0.7) +
  geom_point(color = "purple")
  labs(
    title = "Income vs Unemployment (Bubble = Poverty)"
  )
```

## 6. correlation analysis

## How strongly are poverty indicators associated with grad rates in MONTGOMERY COUNTY?

Correlation analysis

```{r}
MCPS_clean <- MCPS_clean |>
  mutate(
    grad_rate =
      as.numeric(gsub("[^0-9.]", "", grad_rate)),

    FARMS_rate =
      as.numeric(gsub("[^0-9.]", "", FARMS_rate)),

    dropout_rate =
      as.numeric(gsub("[^0-9.]", "", dropout_rate)),

    attendance_rate =
      as.numeric(gsub("[^0-9.]", "", attendance_rate))
  )
```

```{r}
cor(MCPS_clean$FARMS_rate, MCPS_clean$grad_rate, use = "complete.obs")
```

```{r}
MD_clean |> 
  select(poverty_rate, median_income, unemployment_rate) |>
  cor(use = "complete.obs")
```

```{r}
ggplot(MD_clean, aes(x = unemployment_rate, y = poverty_rate)) +
  geom_point(color = "darkred", size = 3) +
  geom_smooth(method = "lm") +
  labs(
    title = "Unemployment vs Poverty (Maryland Counties)"
  ) +
  theme_minimal()
```

```{r}
cor_matrix <- MCPS_clean |>
  select(FARMS_rate, grad_rate, dropout_rate, attendance_rate) |>
  cor(use = "complete.obs")
```

```{r}
cor_df <- as.data.frame(as.table(cor_matrix))

ggplot(cor_df, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = round(Freq, 2)), color = "white") +
  scale_fill_gradient2(low = "darkred",
                       mid = "white",
                       high = "darkblue",
                       midpoint = 0) +
  labs(
    title = "Correlation Matrix of School Variables",
    x = "",
    y = ""
  ) +
  theme_minimal()
```

## 7. Regression models

```{r}
MCPS_model <- MCPS_clean |>
  filter(
    !is.na(grad_rate),
    !is.na(FARMS_rate),
    !is.na(dropout_rate),
    !is.na(attendance_rate)
  )
```

```{r}
final_model <- lm(
  grad_rate ~ FARMS_rate + attendance_rate + dropout_rate,
  data = MCPS_model
)

summary(final_model)
```

```{r}
MCPS_model <- MCPS_model |>
  mutate(
    predicted_grad = predict(final_model)
  )
```

```{r}
ggplot(MCPS_model,
       aes(x = grad_rate,
           y = predicted_grad)) +
  geom_point(color = "darkblue", size = 3) +
  geom_abline(slope = 1,
              intercept = 0,
              color = "red",
              linetype = "dashed") +
  labs(
    title = "Actual vs Predicted Graduation Rates",
    x = "Actual Graduation Rate",
    y = "Predicted Graduation Rate"
  ) +
  theme_minimal()
```

```{r}
summary(final_model)$r.squared
```

```{r}
MCPS_model <- MCPS_model |>
  mutate(
    residual = grad_rate - predicted_grad)
```

```{r}
summary(MCPS_model)
```

## 8. Geographic analysis and maps

## end ?
